Gender Bias
========================================================
Notes
-----
use gc() - garbage Collection -  to free up memory

Prepare Data
------------

### Load Raw Data 

Read in the data:
```{r}
file.location <- "../Data Sets/san diego research file/"

scores.file.name <- "ca2012_all_37_csv_v3.txt"
entities.file.name <- "ca2012entities_csv.txt"
subgroups.file.name <- "Subgroups.txt"
tests.file.name <- "Tests.txt"

scores <- read.csv(paste0(file.location, scores.file.name), header=TRUE)
entities <- read.csv(paste0(file.location, entities.file.name), header=TRUE)
subgroups <- read.csv(paste0(file.location, subgroups.file.name), header=TRUE)
tests <- read.csv(paste0(file.location, tests.file.name), header=TRUE)

```

### Clean Data

Look at district information:

```{r eval=FALSE}
d.code <- as.factor(scores$District.Code)
plot(d.code)

# What district is that big spike?
levels(d.code)[match(max(table(d.code)), table(d.code))]
               
# District 68338 -- San Diego Unified (not surprising)
# I live in Poway Unified, how many records there? Poway Unified == 68296
sum(scores$District.Code == 68296)

# 34113. How many combined?

sum(scores$District.Code == 68296 | scores$District.Code == 68338)
# 212488 seems like a good place to start.
```


Seem to have figured out my merge issues so this is no longer necessary.
```{r eval=FALSE}
# Filter scores to District Codes 68296 or 68338
scores <- scores[scores$District.Code == 68296 | scores$District.Code == 68338,]
```


```{r}
#Percentage.Advanced coming in as a factor, change this to numeric
scores$Percentage.Advanced <- as.numeric(scores$Percentage.Advanced)
```


### Merge Data

Attempting to add columns to scores data set. Starting with 663295 obs. Should be the same when I'm done.

**Add School Names**
```{r}
# Only need entities from San Diego County
entities <- entities[entities$County.Name == "San Diego",]
schools <- unique(entities[,c("School.Code", "School.Name")])

scores <- merge(x = scores, y = schools, by="School.Code", all.x=TRUE)


```

**Add District Names:**  

There is something I don't quite understand where the charter schools have a district code for a more general district but the district name is unique:  
District.Code -- District.Name  
10371         -- San Diego County Office of Education  
10371         -- Literacy First Charter  
  
There seem to be 45 unique district codes. If I filter out records with charter numbers, I get a list of 45 codes and 45 District Names. Let's use this for now.  

```{r}
districts <- unique(entities[entities$Charter.Number == 0, c("District.Code", "District.Name")])
length(unique(districts$District.Code))
length(unique(districts$District.Name))

scores <- merge(x = scores, y = districts, by="District.Code", all.x=TRUE)

```

**Subgroup.ID**

```{r}
subgroups <- subgroups[,-1]
names(subgroups)[1] <- "Subgroup.ID"

scores <- merge(x = scores, y = subgroups, by="Subgroup.ID", all.x=TRUE)
```


**Test.ID**  

```{r}
tests <- tests[,names(tests) != "Test.ID.Num"]
names(tests)[1] <- "Test.Id"

scores <- merge(x = scores, y = tests, by="Test.Id", all.x=TRUE)
```



Other Clean up:
```{r eval=FALSE}
#All SD county so we can drop county code
scores <- scores[,-scores$County.Code]

```

Explore Data
------------

Start by just trying to get a feel for what's there

### Percentage Advanced by subgroup type


```{r fig.width=10, fig.height=12}
# par(mar=c(5,4,4,2)) : default values
par(mar=c(5,20,4,2))
tbl <- aggregate(Percentage.Advanced ~ Subgroup.ID, data=scores, mean, na.action=na.omit)

tbl <- merge(x = tbl, y = subgroups, by="Subgroup.ID", all.x=TRUE)

tbl <- tbl[with(tbl, order(All.Students.1, Percentage.Advanced)), ]

get.color <- function(item) {which(levels(tbl$All.Students.1) == item)}
color.id <- lapply(tbl$All.Students.1, get.color)

barplot(tbl$Percentage.Advanced, names.arg=tbl$All.Students, horiz=TRUE, las=2, cex.names=0.65, 
        col=as.numeric(color.id))

```

I think this may be combining a bunch of types together. In the subgroups table, I have all.students and all.students.1. The latter seems to group for former. I need to be able to distinguish between the groups

```{r}

```








